---
title: "Day 3 - Conditionals & Loops"
author: "Erica Dale"
date: "2022-08-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(palmerpenguins)
```

### `if` statements
back-ticks create code-style appearance

```{r}
burrito <- 2.8

if (burrito > 2) {
  print("I love burritos!")
}
```

#### Example with strings,  using a string match

```{r}
my_ships <- c("Millenium Falcon", "X-wing", "Tie-Fighter", "Death Star")

str_detect(my_ships, pattern = "r")
str_detect(my_ships, pattern = "x") # case sensitive

# Looks within each element, and TRUE/FALSE does this pattern display in any element
# What's actually in a phrase?

phrase <- "I love burritos"
if (str_detect(phrase, "love")) {
  print("I am a burrito fan.")
}
```


#### Example of `%in%` operator

```{r}
view(starwars) # dplyr dataset starwars

tatooine_characters <- starwars |> 
  filter(homeworld == "Tatooine")
view(tatooine_characters)

# Use the %in% operator to check for multiple positive matches, shortcut for "or" statements
tatooine_endor_naboo <- starwars |> 
  filter(homeworld %in% c("Tatooine", "Naboo", "Endor"))
```


### `if-else` statement
```{r}
pika <- 89.1

if(pika > 60) {
  print("mega pika")
} else {
  print("normal pika")
}
```

```{r}
marmot <- 2.8
if(marmot < 0.5) {
  print("this is a small marmot")
} else if (marmot >= 0.5 & marmot < 3) {
  print("this is a medium marmot")
} else {
  print("this is a big marmot")
}
```


### Summary table
#### `group_by() %>% summarize()`
Can use this to make pivot tables

```{r}
view(penguins)

unique(penguins$species) # returns the unique values only, does not group
unique(starwars$eye_color)

penguins |> 
  group_by(species, island) |>   # Organizing column by different species
  summarize(mean_bill_length = mean(bill_length_mm, na.rm = TRUE),
            sd_bill_length = sd(bill_length_mm, na.rm = TRUE),
            n = n()) # the count of each BUT includes NA items ignored above
# Output table has species with subgroups of islands, then data on mean and sd 
```

```{r}
# Find the mean and sd of character height in `starwars` grouped by homeworld

starwars |> 
  group_by(homeworld) |> 
  summarize(height_mean = mean(height, na.rm = TRUE),
            heigh_sd = sd(height, na.rm = TRUE),
            n = n())

# Find maximum and min flipper length, sample size, for each species and gender of penguins
penguins |> 
  group_by(species, sex) |> 
  summarize(max_flipper = max(flipper_length_mm, na.rm = TRUE),
            min_flipper = min(flipper_length_mm, na.rm = TRUE),
            n = n())

```

#### `across()`
```{r}
# Across any columns that satisfy condition, apply these
penguins |> 
  group_by(species) |> 
  summarize(across(where(is.numeric), mean, na.rm = TRUE))

# Can add vector of functions to action c(mean, median, sd)
```


### For loop Cumulative calculations
```{r}
tree_height <- c(1, 2, 6, 10, 14, 20)
# Goal is the add next value to sum of the previous value

for (i in seq_along(tree_height)) {
  val = tree_height[i] + tree_height[i + 1] # Refer to next position
  print(val)
}
```


#### For loop conditional statement
```{r}
# Want to convert ages into diff animal years
# Dog years = (human years * 7), elephant years = (human years *.88), goat years = (human years * 4.7)

species <- c("dog", "elephant", "goat", "dog", "elephant", "goat")
age_human <- c(3, 8, 4, 6, 12, 18, 9) # could also use cbind instead of separate vectors

for (i in seq_along(species)) {
  if (species[i] == "dog") { # single bracket because starting with vector, need [[]] if pulling from matrix/df
    animal_age <- age_human[i] * 7
  } else if (species[i] == "elephant") {
    animal_age <- age_human[i] * 0.88
  } else (species[i] == "goat") {
    animal_age <- age_human * 4.7
  }
  print(animal_age)
}   # Where is the break???
```

```{r}
species <- c("dog", "elephant", "goat", "dog", "elephant", "goat")
age_human <- c(3, 8, 4, 6, 12, 18, 9)
animal_ages <- vector(mode = "numeric", length = length(species))

for (i in seq_along(species)) {
  if (species[i] == "dog") {
    animal_age <- age_human[i] * 7
  } else if (species[i] == "elephant") {
    animal_age <- age_human[i] * 0.88
  } else (species[i] == "goat")
    animal_age <- age_human * 4.7
  animal_ages[i] <- animal_age
}   # Where is the break?
```


#### For loops across data frames
```{r}
mean_mtcars <- vector(mode = "numeric", length = ncol(mtcars))

# Decide which reference for length
dim(mtcars)
seq_along(mtcars)
ncol(mtcars)
# All are equal, all are options

for (i in seq_along(mtcars)) {
  mean_val <- mean(mtcars[[i]], na.rm = TRUE)   # Need double brackets to return vector from df
  mean_mtcars[i] <- mean_val
}
mean_mtcars
```


### Functional Programming
Adding syntactic sugar (apply, across, map_df)

`apply()`
```{r}
# Using same example as above
apply(X = mtcars, MARGIN = 2, FUN = mean)
```

`purrr()`
```{r}
# Using same example as above
map(.x = mtcars, .f = mean)

# want return as a data frame
map_df(.x = mtcars, .f = mean)
```

